---
title: "Tailor-made functions and operations"
author: "Laurent R. Bergé"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: journal
    highlight: haddock
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: no
  pdf_document:
    toc: yes
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{guide_customization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---




# Creating your own operations {#sec_creation}

You can add any arbitraty operation to `string_magic`. You only need to 
use the function `string_magic_register_fun` or `string_magic_register_ops`. 

First you need to create a function that will be applied to a character vector. 
That function must have at least the arguments 'x' and '...'. 
Additionnaly, it can have the arguments: 'argument', 'options', 'group', 'group_flag'.

Second, this function must return a vector or a list of two items: `x` and `group`. The latter
form is only used in functions modifying the length or order of the vector.

Finally, you need to register the function with `string_magic_register_fun` and provide it an
alias. Optionnally, you can provide a list of valid options. 

Ex.1: a new operation adding markdown emphasis.
```{r}
library(stringmagic)
# A) define the function
fun_emph = function(x, ...) paste0("*", x, "*")
 
# B) register it
string_magic_register_fun(fun_emph, "emph")

# C) use it
x = string_vec("right, now")
string_magic("Take heed, {emph, c? x}.")
```

Above, we have created a simple functon that adds markdown emphasis to words.

More generally, the function taken by `string_magic_register_fun` is called internally 
by `string_magic` in the form `fun(x, argument, options, group, group_flag)`.
Here is the meaning of the arguments:

- `x`: the value to which the operation applies. 
- `argument`: the quoted `string_magic` argument (always character). 
- `options`: a character vector of `string_magic` options. 
- `group`: an index of the group to which belongs each observation (integer). 
- `group_flag`: value between 0 and 2; 0: no grouping operation requested; 
1: keep track of groups; 2: apply grouping.

The two last arguments, `group` and `group_flag`, are of use
only in group-wise operations only if `fun` changes the length or the order of vectors. 

Let's add an argument and an option to the `"emph"` operation that we defined in Ex.1.

Ex.2: new operation with argument and option.
```{r}
fun_emph = function(x, argument, options, ...){
  arg = argument
  if(nchar(arg) == 0) arg = "*"
  
  if("strong" %in% options){
    arg = paste0(rep(arg, 3), collapse = "")
  }
  
  paste0(arg, x, arg)
}

string_magic_register_fun(fun_emph, "emph", "strong")

x = string_vec("right, now")
string_magic("Take heed, {'_'emph.s, c? x}.")

# In string_magic_register_fun, the valid_option argument is used to validate them.
try(string_magic("Take heed, {'_'emph.aaa, c? x}."))
```

Finally let's illustrate an example with group-wise axareness. 

Ex.3: we create a function that only keeps variable names (ex: x5, is_num, etc).
```{r}

keep_varnames = function(x, group, group_flag, ...){
  is_ok = grepl("^[[:alpha:].][[:alnum:]._]*$", x)
  
  if(group_flag != 0){
    group = group[is_ok]
    group = stringmagic:::cpp_recreate_index(group)
  } 
  
  res = list(x = x[is_ok], group = group)
  
  return(res)
}

string_magic_register_fun(keep_varnames, "keepvar")

expr = c("x1 + 52", "73 %% 5 == x", "y[y > .z_5]")
string_magic("All vars: {'[^[:alnum:]_.]+'s, keepvar, unik, bq, C ? expr}.")

# thanks to the group flag, we can apply group-wise operations
string_magic("Vars in each expr:\n",
        "{'\n'c ! - {1:3}) {'[^[:alnum:]_.]+'s, keepvar, bq, ~(unik, C) ? expr}}")
```
