---
title: "`string_magic`'s operations: The reference"
author: "Laurent R. Bergé"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: journal
    highlight: haddock
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: no
  pdf_document:
    toc: yes
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{operations_reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(stringmagic)
```

This document references all regular `string_magic` operations. They can be used within 
`string_magic` or can be accessed from `string_ops` or `string_clean`. 

The operations are divided into four groups:

- [basic string operations](#sec_basic)
- [operations changing the length or order](#sec_op_length)
- [formatting operations](#sec_format)
- [other operations](#sec_other)

By default the function `string_magic` returns a plain character vector. In this 
vignette it is sometimes nicer to apply the function `base::cat` to display 
`string_magic` results containing newlines. Ths function `cat_magic` does exactly 
that and we will use it from time to time.


# Basic string operations {#sec_basic}

This section describes some of the most common string operations: extracting, 
replacing, collapsing, splitting, etc. Because they are so common, many of these 
operations have a one letter alias.
These functions accept regex flags in their patterns. For more information on regex flags, 
see the [dedicated section](https://lrberge.github.io/stringmagic/articles/guide_string_tools.html).

## Split strings: s, S, split, Split: 

Splits the string according to a pattern. The four operations have different defaults: `' '` 
for `s` and `split`, and `',[ \t\n]*'` for `S` and `Split` (i.e. comma separation). 

When character strings are split, their identity is kept in memory 
so that group-wise operations can be applied. See the [section on group-wise operations](https://lrberge.github.io/stringmagic/articles/ref_string_magic_special_operations.html#sec_group_wise).
```{r}
# 'Split' with its default (comma separation)
string_magic("{Split ! romeo, juliet}")

# result with 's' is different
string_magic("{split ! romeo, juliet}")

# with argument: 's' and 'S' are identical
# note the flag 'fixed' (`f/`) to remove regex interpretation
string_magic("{'f/+'split, '-'collapse ! 5 + 2} = 3")

# group wise operations (here `~(sort, collapse)`, see dedicated section)
prince_talk = c("O that this too too solid flesh would melt",
                "Thaw, and resolve itself into a dew!",
                "Or that the Everlasting had not fix'd",
                "His canon 'gainst self-slaughter!")
cat_magic("Order matters:\n{split, ~(sort, collapse), align.center, lower, upper.sentence,
                            Q, '\n'collapse ? prince_talk}")
```

## Collapse strings: c, C, collapse, Collapse

To collapse (or concatenate) multiple strings into a single one. The four operations are 
identical, only their defaults change. The default is `' '` for `c` and `collapse`, 
and `', | and '` for `C` and `Collapse`.
The syntax of the argument is `'s1'` or `'s1|s2'`. `s1` is the string used to concatenate 
(think `paste(x, collapse = s1)`). In arguments of the form `'s1|s2'`, `s2` will be 
used to concatenate the last two elements. 

```{r}
# regular way
x = 1:4
string_magic("And {' and 'collapse ? x}!")

# with s2
string_magic("Choose: {', | or 'collapse ? 2:4}?")

# default of Collapse: enumeration (similar to operation enum)
wines = c("Saint-Estephe", "Margaux")
string_magic("I like {Collapse ? wines}.")

# default of collapse: space concatenation
string_magic("{split, '.{5,}'get, collapse ! I don't like short words}")
```

## Extract patterns: extract, x, X

Extracts the first or multiple patterns from a string. Default argument is `'[[:alnum:]]+'`.
Command `"extract"` accepts the option `"first"`, and 
`"x"` and `"X"` accept no option. `x` is an alias for `extract.first` and `X` for `extract`.
Use the option `"first"` to extract only the first match for each string.

When patterns are extracted, the identity of each original character string is kept in memory 
so that group-wise operations can be applied. See the [section on group-wise operations](https://lrberge.github.io/stringmagic/articles/ref_string_magic_special_operations.html#sec_group_wise).
```{r}
x = c("margo: 32, 1m75", "luke doe: 27, 1m71")
string_magic("{'^\\w+'extract ? x} is {'\\d+'extract.first ? x}")

# illustrating multiple extractions
# group-wise operation (~()) is detailed in its own section
x = c("Combien de marins, combien de capitaines.",
      "Qui sont partis joyeux pour des courses lointaines,",
      "Dans ce morne horizon se sont évanouis !")
string_magic("Endings with i: {'i\\w*'extract, ~(', 'collapse), enum.1 ? x}.")


x = c("6 feet under", "mahogany")
# single extraction
string_magic("{'\\w{3}'x ? x}")

# multiple extraction
string_magic("{'\\w{3}'X ? x}")
```

## Replace patterns: r, R, replace

Replaces a pattern with a string. The three operators are identical and have no default.
The syntax is `'flags/old'` or `'old => new'` with `'old'` the pattern to find and `new` the 
replacement. `flags/` are optional regex flags. The default for `new` is 
the empty string. On top of regular regex flags, this operation also accepts 
the flags `"total"` and `single`. `total` instructs to replace the fulll string in case 
the pattern is found.

The flag `"single"` leads to only a single substitution per string (the first pattern is replaced). 
That is, the function `base::sub` is used instead of `base::gsub`.
```{r}
# regex without replacement (ie removing)
string_magic("{'e'replace ! Where is the letter e?}")

# regex with replacement
string_magic("{'(?<!\\b)e => a'replace ! Where is the letter e?}")

# with option "single"
string_magic("{'single/(?<!\\b)e => a'replace ! Where is the letter e?}")

# we replace the full string with the flag total (`t/`)
x = c("Where is the letter e?", "Not this way!")
string_magic("{'total/e => here!'r ? x}")
```

## Clean string patterns: clean {#op_clean}

Replaces a pattern with a string. Similar to the operation `r`, except that here the comma is
a pattern separator. The argument is of the form `"flags/pattern1, pattern2 => replacement"`. 
See detailed explanations in [`string_clean()`](https://lrberge.github.io/stringmagic/articles/guide_string_tools.html#sec_clean). 

```{r}
# we use the fixed pattern to remove the regex meaning
string_magic("{'f/[, ]'clean ! x[a]}")
```

## Get selected strings: get

Restricts the string vector to only the values respecting a pattern. This operation has no default.
Accepts the options `"equal"` and `"in"`.
By default it uses the same syntax as [`string_get()`](https://lrberge.github.io/stringmagic/articles/guide_string_tools.html#detect_funs) 
so that you can use regex flags and 
include logical operations between regex patterns with `' & '` and `' | '`.
If the option `"equal"` is used, a simple string equality with the argument is tested (hence
no flags are accepted). If the option `"in"` is used, the argument is first split with respect to commas
and then set inclusion is tested. 

```{r}
x = row.names(mtcars)
#  we only keep models containing "Merc" and ending with a letter ([[:alpha:]]$)
string_magic("Mercedes models: {'Merc & [[:alpha:]]$'get, '^.+ 'r, enum ? x}.")

models = c("Merc 230", "Merc 450SE", "Merc 480")
# we only ekep the ones in the set
string_magic("Mercedes models: {`models`get.in, enum ? x}.")
```

## Detect patterns in strings: is

Detects if a pattern is present in a string, returns a logical vector. This operation has no default.
Mostly useful as the final operation in a [`string_ops()`](https://lrberge.github.io/stringmagic/articles/guide_string_tools.html#sec_ops) call.
By default it uses the same syntax as [`string_is()`](https://lrberge.github.io/stringmagic/articles/guide_string_tools.html#detect_funs) 
so that you can use regex flags and 
include logical operations between regex patterns with `' & '` and `' | '`.
If the option `"equal"` is used, a simple string equality with the argument is tested (hence
no flags are accepted). If the option `"in"` is used, the argument is first split with respect to commas
and then set inclusion is tested. 
```{r}
x = c("Mark", "Lucas")
# note that we use the flag `i/` to ignore the case
string_magic("Mark? {'i/mark'is, enum ? x}")
```

## Get the index of the strings containing a pattern: which

Returns the index of string containing a specified pattern. With no default, can be applied
to a logical vector directly. 
By default it uses the same syntax as [`string_which()`](https://lrberge.github.io/stringmagic/articles/guide_string_tools.html#detect_funs) 
so that you can use regex flags and 
include logical operations between regex patterns with `' & '` and `' | '`.
If the option `"equal"` is used, a simple string equality with the argument is tested (hence
no flags are accepted). If the option `"in"` is used, the argument is first split with respect to commas
and then set inclusion is tested. 
Mostly useful as the final operation in a [`string_ops()`](https://lrberge.github.io/stringmagic/articles/guide_string_tools.html#sec_ops) call.
```{r}
x = c("Mark", "Lucas", "Markus")
# note that we use the flag `i` to ignore the case and `w` to add word boundaries
string_magic("Mark is number {'iw/mark'which ? x}.")
```

# Operations changing the length or the order {#sec_op_length}

## Keep only the first elements: first

Keeps only the first `n` elements. 

```{r}
string_magic("First 3 mpg values: {3 first, enum ? mtcars$mpg}.")

# you could have done the same with regular R in the expression...
string_magic("First 3 mpg values: {enum ? head(mtcars$mpg, 3)}.")

# ...but not in the middle of an operations chain
string_magic("First 3 integer mpg values: {'f/!.'get, 3 first, enum ? mtcars$mpg}.")
```

Negative numbers as argument remove the enum
first `n` values. You can add a second argument in the form `'n1|n2'first` in which case the first `n1` and last
`n2` values are kept; `n1` and `n2` must be positive numbers.

```{r}
string_magic("Letters in the middle: {13 first, 5 last, enum ? letters}.")

string_magic("First and last letters: {'3|3'first, enum ? letters}.")

string_magic("Last letters: {-21 first, enum ? letters}.")
```

## Keep only the first elements (alternative): K

Keeps only the first `n` elements; has more options than `first`. The syntax is `'n'K`, 
`'n|s'K`, `'n||s'K`. `n` provides the number of elements to keep. If `s` is provided and the number of 
elements are greater than `n`, then in `'n|s'` the string `s` is added at the end, and
if `'n||s'` the string `s` replaces the nth element.
The string `s` accepts specials values:
  + `:n:` or `:N:` which gives the total number of items in digits or letters (N)
  + `:rest:` or `:REST:` which gives the number of elements that have been truncated in digits or letters (REST)

```{r}
# basic use
string_magic("First 3 letters: {'3'K, q, enum ? letters}.")

# advanced use: using the extra argument
string_magic("The letters are: {q, '3|:rest: others'K, enum ? letters}.")
```

## Keep only the last elements: last

Keeps only the last `n` elements. Negative numbers as argument remove the 
last `n` values.
```{r}
string_magic("Last 3 mpg values: {3 last, enum ? mtcars$mpg}.")

string_magic("Removing the 3 last elements leads to {-3 last, enum ! x{1:5}}.")
```

## Sort the vector: sort

Sorts the vector in increasing order. Accepts an optional argument and the option `"num"`. 
```{r}
x = c("sort", "me")
# basic use
string_magic("{sort, collapse ? x}")
```
If an argument is provided, it must be a regex pattern that will be applied to
the vector using [`string_clean()`](https://lrberge.github.io/stringmagic/articles/guide_string_tools.html#sec_clean). 
The sorting will be applied to the modified version of the vector
and the original vector will be ordered according to this sorting. 
```{r}
# first modifying the string before sorting
# here the regex first removes the first word, meaning that we sort on the last names
x = c("Jon Snow", "Khal Drogo")
string_magic("{'.+ 'sort, enum?x}")
```

The option `"num"` sorts over a numeric version 
(with silent conversion) of the vector and reorders the original vector accordingly. 
Values which could not be converted are last.
```{r}
x = "Mark is 34, Bianca is 55, Odette is 101, Julie is 21 and Frank is 5"
# sort on the "character string" number
string_magic("{', | and 'split, '\\D'sort, enum ? x}")

# we extract the numbers, then convert to numeric, then sort
string_magic("{', | and 'split, '\\D'sort.num, enum ? x}")
```
**Important note**: the sorting operation is applied before any character conversion.
If previous operations were applied, it is likely that numeric data were transformed to character.
```{r}
# note the difference
x = c(20, 100, 10)

# sorting on numeric
string_magic("{sort, ' + 'collapse ? x}")

# sorting on character since 'n' operation transformed the vector to character
string_magic("{n, sort, ' + 'collapse ? x}")
```

## Sort the vector in decreasing order: dsort

Sorts the vector in decreasing order. It accepts an optional argument and 
the option `"num"`. See the operation `"sort"` for a description of the argument and the option.
```{r}
string_magic("5 = {dsort, ' + 'collapse ? 2:3}")
```

## Reverse the vector: rev

Reverses the vector.
```{r}
string_magic("{rev, ''collapse ? 1:3}")
```

## Keep only unique elements: unik

Makes the string vector unique. 
```{r}
string_magic("Iris species: {unik, upper.first, enum ? iris$Species}.")
```

## Repeat each elements of the vector: each

Repeats each element of the vector `n` times. Option `"c"` then collapses the full vector 
with the empty string as a separator. 
```{r}
# note: operation `S` splits splits wrt to commas (default behavior)
string_magic("{S!x, y}{2 each ? 1:2}")

# illustrating collapsing
string_magic("Large number: 1{5 each.c ! 0}")
```

## Repeats the vector: times

Repeats the vector sequence `n` times. Option `"c"` then collapses the full vector 
with the empty string as a separator. 
```{r}
string_magic("What{6 times.c ! ?}")
```

## Remove specific values: rm

Removes elements from the vector. Options: `"empty"`, `"blank"`, `"noalpha"`, `"noalnum"`, `"all"`.
The *optional* argument represents the pattern used to detect strings to be deleted. 
```{r}
x = c("Luke", "Charles")
string_magic("{'i/lu'rm ? x}")
```

By default it removes empty strings. 
Option `"blank"` removes strings containing only blank characters (spaces, tab, newline).
Option `"noalpha"` removes strings not containing letters. Option `"noalnum"` removes strings not 
containing alpha numeric characters. Option `"all"` removes all strings (useful in conditions, see 
the dedicated section). If an argument is provided, only the options `"empty"` and `"blank"` are available.

```{r}
x = c("I want to enter.", "Age?", "21.")
string_magic("Nightclub conversation: {rm.noalpha, c ! - {x}}")
```

## Remove all value: nuke

Removes all elements, equivalent to `rm.all` but possibly more explicit. 
Useful in conditions, see the dedicated section.
```{r}
x = c(5, 7, 453, 647)
# here we use a condition: see the dedicated section for more information
string_magic("Small numbers only: {if(.>20 ; nuke), enum ? x}.")
```

## Insert a character string: insert

Inserts a new element to the vector. Options: `"right"` and `"both"`. Option `"right"` adds
the new element to the right. Option `"both"` inserts the new element on the two sides of the vector.

```{r}
string_magic("{'3'insert.right, ' + 'collapse ? 1:2}")
```

# Formatting operations {#sec_format}
 
## Change the case: lower

Lower cases the full string.
```{r}
x = "MesSeD uP CaSe"
string_magic("from a {x} to {lower?x}")
```

## Change the case: upper

Upper cases the full string. Options: `"first"` and `"sentence"`.
Option `"first"` upper cases only the first character. Option `"sentence"`
upper cases the first letter after punctuation. 

```{r}
x = "hi. how are you? fine."
string_magic("{upper.sentence ? x}")
```

## Change the case: title

Applies a title case to the string. Options: `"force"` and `"ignore"`.
Option `"force"` first puts everything to lowercase before applying the title case. 
Option `"ignore"` ignores a few small prepositions ("a", "the", "of", etc).

```{r}
x = "bryan is in the KITCHEN"

# default: respects upper cases
string_magic("{title ? x}")

# force: force to title case
string_magic("{title.force ? x}")

# ignore: ignores small prepositions
string_magic("{title.force.ignore ? x}")
```

## Normalize white spaces: ws

Normalizes whitespaces (WS). It trims the whitespaces on the edges and transforms any succession 
of whitespaces into a single one. Can also be used to further clean the string with its options. 
Options: `"punct"`, `"digit"`, `"isolated"`. Option `"punct"` cleans the punctuation. Option `"digit"` cleans digits.
Option `"isolated"` cleans isolated letters. WS normalization always come after any of these options.
**Important note:** punctuation (or digits) are replaced with WS and **not** 
the empty string. This means that `string_magic("ws.punct ! Meg's car")` will become `"Meg s car"`.

```{r}
x = "   I    should? review 85 4 this text!!"
cat_magic("v0: {x}", 
          "v1: {ws ? x}",
          "v2: {ws.punct ? x}",
          "v3: {ws.punct.digit ? x}",
          "v4: {ws.punct.digit.isolated ? x}", .sep = "\n")
```

## Trim white spaces: tws

Trims the white spaces on both ends of the strings.
```{r}
x = "  too much space \t\n"
string_magic("With trim: {tws, Q ? x}")
```

## Add various type of quotes: q, Q, bq

To add quotes to the strings. q: single quotes, Q: double quotes, bq: 
back quotes.
```{r}
x = c("Mark", "Pam")
string_magic("Hello {q, enum ? x}!")
```

## Format the values with `base::format`: format, Format

Applies the base R's function `base::format()` to the string. 
By default, the values are left aligned, *even numbers* (differently from `base::format()`'s behavior).
The upper case command (`Format`) applies right alignment. Options: `"0"`, `"zero"`, `"right"`, `"center"`.
Options `"0"` or `"zero"` fills the blanks with 0s: useful to format numbers. Option `"right"` right aligns,
and `"center"` centers the strings.
```{r}
x = c(1, 12345) 
cat_magic("left  : {format, q, enum ? x}", 
          "right : {Format, q, enum ? x}",
          "center: {format.center, q, enum ? x}",
          "zero  : {format.0, q, enum ? x}", .sep = "\n")
```

## Apply `sprintf` formatting: %

Applies `base::sprintf()` formatting. The syntax is `'arg'%` with arg an sprintf formatting,
or directly the sprint formatting.
```{r}
string_magic("pi = {%.3f ? pi}")
```

## Remove stop words: stopwords

Removes basic English stopwords (the snowball list is used). 
The stopwords are replaced with an empty space but the left and right WS are 
untouched. So WS normalization may be needed (see operation `ws`).
```{r}
x = c("He is tall", "He isn't young")
string_magic("Is he {stop, ws, enum ? x}?")
```

## Turn the string to ASCII: ascii

Turns all letters into ASCII with transliteration. Failed translations are transformed 
into question marks. Options: `"silent"`, `"utf8"`. By default, if some conversion fails
a warning is prompted. Option `"silent"` disables the warning in case of failed conversion. The conversion 
is done with `base::iconv()`, option `"utf8"` indicates that the source endocing is UTF-8, can be useful 
in some cases.
```{r}
author = "Laurent Bergé"
string_magic("This package has been developped by {ascii ? author}.")
```

## Formatting integers: n, N

Formats integers by adding a comma to separate thousands. Options: `"letter"`, `"upper"`, 
`"0"`, `"zero"`, `"roman"`, `"Roman"`.
The option `"letter"` writes the number in letters (large numbers keep their numeric format). The option
`"upper"` is like the option `"letter"` but uppercases the first letter. Options `"0"` or `"zero"` left pads
numeric vectors with 0s. The options `"roman"` and `"Roman"` write the integer in Roman 
with `utils::as.roman`. The lower case version writes them in lower case. 
The upper case command (`N`) adds the option `"letter"`.
```{r}
x = c(5, 12, 52123)
string_magic("She owes {n, '$'paste, enum ? x}.")

# option 0: all same width, no ',' for thousands
cat_magic("|---|\n{n.0, '\n'collapse ? x}")

# option "upper"
n = 5
string_magic("{n.upper ? n} is my favourite number.")

# N: like "n.letter"
x = 5
string_magic("He's {N ? x} years old.")

# roman
string_magic("What's nicer? {collapse?11:13}, {n.roman, c?11:13} or {n.Roman, c?11:13}?")
```

## Numbered position: nth, Nth

When applied to a number, this operator writes them as a rank. Options: `"letter"`, 
`"upper"`, `"compact"`. 
```{r}
n = c(3, 7)
string_magic("They finished {nth, enum ? n}!")
```
Option `"letter"` tries to write the numbers in letters, but note that it stops at 20. Option `"upper"`
is the same as `"letter"` but uppercases the first letter. Option `"compact"` aggregates
consecutive sequences in the form `"start_n_th to end_n_th"`. 
```{r}
string_magic("They arrived {nth.compact ? 5:20}.")
```
The upper case command (`Nth`) adds the option `"letter"`.
```{r}
n = c(3, 7)
string_magic("They finished {Nth, enum ? n}!")
```

## Number of times: ntimes, Ntimes

Write numbers in the form `n` times. Options: `"letter"`, `"upper"`. Option 
`"letter"` writes the number in letters (up to 100). Option `"upper"` does the same as `"letter"` 
and uppercases the first letter. 
```{r}
string_magic("They lost {enum ! {ntimes ? c(1, 12)} against {S!Real, Barcelona}}.")
```
The upper case command (`Ntimes`) adds the option `"letter"`.
```{r}
x = 5
string_magic("This paper was rejected {Ntimes ? x}...")
```

## Keep only the first, last characters: firstchar, lastchar

To select the first/last characters of each element. 
Negative numbers remove the first/last characters.
```{r}
string_magic("{19 firstchar, 9 lastchar ! This is a very long sentence}")

string_magic("delete 3 = {-3 firstchar ! delete 3}")
```

## Shortens character strings: k, shorten, Shorten

To keep only the first `n` characters (like `firstchar` but with more options). 
(Note that `k` stands for "keep" and exists for historical reasons.)
Available options: `"include"`, `"dots"`. The argument can be of the form `'n'` or `'n|s'` 
with `n` a number and `s` a string. `n` provides the number of characters to keep. 
Optionnaly, only for strings *whose length is greater than `n`*, 
after truncation, the string `s` is appended at the end.

By default, if argument `s` is provided, strings longer than `n` end up at size `n + nchar(s)`.
If option `"include"` is provided, the strings are guaranteed to be of maximum size `n`,
even after the string `s` has been appended. Example: if `n=4` and `s=".."`, then "hello"
becomes "hell.." without `"include"`, and "he.." with it.

Option `"dots"`: if strings are longer than `n+1`, they are truncated at `n-1` and two dots are 
appended. For example if `n = 3`, then "hello" becomes "he..". Disregards the argument `s`.
The operation `"Shorten"` (upper case), is with the option `"dots"`.

Note that another way to add the option `"include"` is to use a double pipe
for the argument `s`, like in `'n||s'`.

```{r}
x = "long sentence"
cat_magic("v0: {x}", 
          "v1: {4 shorten ? x}", 
          "v2: {'4|..'shorten ? x}", 
          "v3: {'4|..'shorten.include ? x}", 
          "v4: {4 shorten.dots ? x}", .sep = "\n")
```

## Fill character strings: fill, align

Fills the character strings up to a size. `align` is an alias to `fill`. 
Options: `"right"` or `"center"`. Default is left-alignment of the strings. 

The argument is optional and can be of the form `'n'` or `'n|s'`. By default if no argument
is provided, of if `n=0`,`n` is equal to the maximum character length of the vector. The optional
argument `s` is a symbol used to fill the blanks. By default `s` is equal to a white space.

Option `"right"` right aligns and `"center"` centers the strings.

See help for [`string_fill()`](https://lrberge.github.io/stringmagic/reference/string_fill.html) 
for more information.
```{r}
life = "full of sound and fury, Signifying nothing"
cat_magic("{'[ ,]+'split, upper.first, fill.center, q, '\n'collapse ? life}")

# fixing the length and filling with 0s
string_magic("{'5|0'fill.right, enum ? c(1, 55)}")
```

## Append text: paste and append

Pastes a custom character string to all elements of the string. The operations `paste` and `append` are equivalent.
This operation has no default. 
Options: `"both"`, `"right"`, `"front"`, `"back"`, `"delete"`. By default, a string is pasted on the left.
Option `"right"` pastes on the right and `"both"` pastes on both sides. Option `"front"` only 
pastes on the first element while option `"back"` only pastes on the last element. Option `"delete"`
first replaces all elements with the empty string.
```{r}
string_magic("y = {'x'paste, ' + 'collapse ? 1:3}")
```

The argument can be of the form `s1` or `s1|s2`. If of the second form, this is equivalent 
to chaining two `paste` operations, once on the left and once on the right: `'s1'paste, 's2'paste.right`.

```{r}
string_magic("y = {'x|0'paste, ' + 'collapse ? 1:3}")
```

## Join lines: join

Joins lines ending with a double backslash. 

```{r}
sun = "The sun \\
is shining."
string_magic("How is the sun? {join ? sun}")
```

## Escape special characters: escape

Adds backslashes in front of specific characters. Options `"nl"`, `"tab"`. 
Option `"nl"` escapes the newlines (`\n`), leading them to be displayed as `"\\\\n"`.
Option `"tab"` does the same for tabs (`"\t"`). This is useful to make the value free
of space formatters. 
The default behavior is to escape both newlines and tabs.

```{r}
input = "yes \n\n no"
msg = string_magic("Your input, equal to {escape, bq ? input} is incorrect.")
cat(msg, sep = "\n")
```


# Other operations {#sec_other}

## Convert to numeric: num

Converts to numeric. Options: `"warn"`, `"soft"`, `"rm"`, `"clear"`. By default, the conversion
is performed silently and elements that fail to convert are turned into NA. 
Option `"warn"` displays a warning if the conversion to numeric fails. 
Option `"soft"` does not convert if the conversion of at least one element fails, leading to 
a character vector. 
Option `"rm"` converts and removes the elements that could not be converted. 
Option `"clear"` turns failed conversions into the empty string, and hence lead to a character vector.
```{r}
x = c(5, "six")
cat_magic("   origin: {enum, q ? x}", 
          "      num: {num, enum, q ? x}", 
          "   num.rm: {num.rm, enum, q ? x}", 
          " num.soft: {num.soft, enum, q ? x}", 
          "num.clear: {num.clear, enum, q ? x}", .sep = "\n")
```

## Create an enumeration: enum

Enumerates the elements. It creates a single string containing the comma 
separated list of elements.
If there are more than 7 elements, only the first 6 are shown and the number of
items left is written.
```{r}
string_magic("{enum ? 1:5}")
```
You can add the following options:

  + `q`, `Q`, or `bq`: to quote the elements
  + `or`, `nor`: to finish with an 'or' (or 'nor') instead of an 'and'
  + `i`, `I`, `a`, `A`, `1`: to enumerate with this prefix, like in: i) one, and ii) two
  + a number: to tell the number of items to display
```{r}
x = c("Marv", "Nancy")
string_magic("The murderer must be {enum.or ? x}.")

x = c("oranges", "milk", "rice")
string_magic("Shopping list: {enum.i.q ? x}.")

# enum is made for display: when vectors are too long, they are truncated
# default is at 7
x = string_magic("x{sample(100, 30)}")
string_magic("The problematic variables are {'x'sort.num, enum ? x}.")

# you can augment or reduce the numbers to display with an option
string_magic("The problematic variables are {'x'sort.num, enum.3 ? x}.")
```


## Formatted length: len, Len

Gives the length of the vector. Options `"letter"`, `"upper"`, `"num"`.
Option `"letter"` writes the length in words (up to 100). Option `"upper"` is the same 
as letter but uppercases the first letter. By default, the number is formatted:
commas are inserted to separate thousands.
```{r}
cat_magic("The length of 1:5000:", 
          " - len     = {len ? 1:5000}", 
          " - len.num = {len.num ? 1:5000}", .sep = " \n")
```
The upper case command (`Len`) adds the option `"letter"` (only for small numbers).
```{r}
string_magic("Its size is {Len ? 1:8}")
```

## Add newlines to force the string to fit a given width: width

Formats the string to fit a given width by cutting at word boundaries. 
Accepts arguments of the form `'n'` or `'n|s'`, with `n` a number and `s` a string. 
An argument of the form `'n|s'` will add `s` at the beginning of each line. Further,
by default a trailing white space is added to `s`; to remove this 
behavior, add an underscore at the end of it. 
The argument `n` is either 
an integer giving the target character width (minimum is 15), or it can be a fraction expressing the 
target size as a fraction of the current screen. Finally it can be an expression that 
uses the variable `.sw` which will capture the value of the current screen width.
```{r}
x = "this is a long sentence"
cat_magic("------ version 0 ------\n{x}", 
          "------ version 1 ------\n{15 width ? x}", 
          "------ version 2 ------\n{'15|#>'width ? x}",
          "------ version 3 ------\n{'15|#>_'width ? x}", .sep = "\n")
```

## Formatted time difference: difftime

Displays a formatted time difference. Option `"silent"` does not report a warning if the
operation fails. It accepts either objects of class `POSIXt` or `difftime`.
```{r}
x = Sys.time()
Sys.sleep(0.15) 
string_magic("Time: {difftime ? x}")
```



