---
title: "String tools: Magick edition"
author: "Laurent R. Bergé"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: journal
    highlight: haddock
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: no
      smooth_scroll: no
  pdf_document:
    toc: yes
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{string_tools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# we preload it to avoid ugly (was comiled with R.x.x) warnings in the doc
library(stringmagick)

options("smagick_str_get_forced_caching" = TRUE)
```

This vignette describes `stringmagick` tools for handling character vectors. 
For a quick start, here are the sections describing:

- how to [detect complex regex (=regular expression) patterns](#detect_funs)
- how to easily, and clearly, [chain multiple string operations](#sec_ops)
- how to efficiently, and clearly, [clean character strings](#sec_clean)

This vignette is also a reference for i) the [optional regex flags](#sec_regex) available to all
`stringmagick` regular expressions, and ii) the [logical combination of regular expressions](#detect_logic).


# Regex flags {#sec_regex}

All functions in `stringmagick` accept optional regex flags when regular expressions
are expected. The idea is similar to 
[regular regex flags](https://developer.mozilla.org/en-US/docs/Web/JavaScript/Guide/Regular_expressions#advanced_searching_with_flags),
but the flags are different in name and effect.

## Regex flags: Syntax

Use `"flag1, flag2/regex"` to add the flags `flag1` and `flag2` to the regular expression `regex`.
For example `"ignore, fixed/dt["` will add the flags `ignore` and `fixed` to the regex `dt[`.

Alternatively, use only the initials of the flags. Hence, `"if/dt["` would also add the 
flags `ignore` and `fixed`.

If the regex does not contain a slash (`/`), no flags are added. If your regex should 
contain a slash, see the [section on escaping](#flag_escaping).

Ex: let's find lines containing `"dt["`:
```{r}
code = c("DT = as.data.table(iris)", 
         "DT[, .(pl_sl = smagick('PL/SL = {Petal.Length / Sepal.Length}')]")

str_get(code, "if/dt[")
```

## Regex flags: Reference

The available flags are:

- [ignore](#flag_ignore): always available
- [fixed](#flag_fixed): always available
- [word](#flag_word): always available
- [magic](#flag_magic): always available
- [total](#flag_total): only available in functions performing a replacement
- [single](#flag_single): only available in functions performing a replacement

### ignore {#flag_ignore}

The flag `"ignore"` leads to a case-insensitive search. 

Ex: let's extract words starting with the last letters of the alphabet.
```{r}
unhappy = "Rumble thy bellyful! Spit, fire! spout, rain!
Nor rain, wind, thunder, fire are my daughters.
I tax not you, you elements, with unkindness.
I never gave you kingdom, call'd you children,
You owe me no subscription. Then let fall
Your horrible pleasure. Here I stand your slave,
A poor, infirm, weak, and despis'd old man."

# the ignore flag allows to retain words starting with the
# upper cased letters
# ex: getting words starting with the letter 'r' to 'z'
smagick("{'i/\\b[r-z]\\w+'extract, c, 60 width ? unhappy}")
```

*Technically*, the [perl](https://www.pcre.org/) expression `"(?i)"` is added at the 
beginning of the pattern.

### fixed {#flag_fixed}

The flag `"fixed"` removes any special regular expression meaning from the pattern, 
and treats it as verbatim.

Ex: let's fix the equation by changing the operators. 
```{r}
x = "50 + 5 * 5 = 40"
str_clean(x, "f/+, * => -")

# Without the fixed flag, we would have gotten an error since '+' or '*'
# have a special meaning in regular expressions and expects something before

# Here's the error
try(str_clean(x, "+, * => -"))
```

*Technically*, if `"fixed"` is the only flag, then the functions `base::grepl` or `base::gsub`
are run with the argument `fixed = TRUE`. If there are also the flags `"ignore"` or `"word"`, 
the pattern is nested into the perl boundaries `\\Q` and `\\E` which strip any special meaning
from the pattern.

### word {#flag_word}

The flag `"word"`:

- adds word boundaries to the pattern
- accepts comma-separated enumerations of words which are concatenated with a logical 'or'

The logic of accepting comma-separated enumerations is to increase readability.
For example, with the flag `"word"`, `"is, are, were"` is equivalent to `"\\b(is|are|were)\\b"`.

Ex: we hide a few words from Alfred de Vigny's poem.
```{r}
le_mont_des_oliviers = "S'il est vrai qu'au Jardin sacré des Écritures,
Le Fils de l'homme ai dit ce qu'on voit rapporté ;
Muet, aveugle et sourd au cri des créatures,
Si le Ciel nous laissa comme un monde avorté,
Alors le Juste opposera le dédain à l'absence
Et ne répondra plus que par un froid silence
Au silence éternel de la Divinité."

# we hide a few words from this poem
smagick("{'wi/et, le, il, au, des?, ce => _'r ? le_mont_des_oliviers}")
```

*Technically*, first the pattern is split with respect to `",[ \t\n]+"`, then all elements
are collapsed with `"|"`. If the flag `"fixed"` was also present, each element is first wrapped 
into `"\\Q"` and `"\\E"`. Finally, we add parentheses (to enable capture) and word 
boundaries (`"\\b"`) on both sides.

### magic {#flag_magic}

The flag `"magic"` allows the user to interpolate variables inside the regular expression
before the regex is evaluated.

Ex: interpolating variables inside regular expressions.
```{r}
vowels ="aeiouy"
# let's keep only the vowels
# we want the pattern: "[^aeiouy]"
lmb = "'Tis safer to be that which we destroy
Than by destruction dwell in doubtful joy."
str_clean(lmb, "magic/[^{vowels}] => _")

#
# Illustration of `smagick` operations before regex application

cars = row.names(mtcars)
# Which of these models contain a digit?
models = c("Toyota", "Hornet", "Porsche")
# we want the pattern "(Toyota|Hornet|"Porsche).+\\d"
# we collapse the models with a pipe using '|'c
str_get(cars, "m/({'|'c?models}).+\\d")

# alternative: same as above but we first comma-split the vector
models_comma = "Toyota, Hornet, Porsche"
str_get(cars, "m/({S, '|'c?models_comma}).+\\d")

#
# Interpolation does not apply to regex-specific curly brackets

# We delete only successions of 2+ vowels
# {2,} has a rexex meaning and is not interpolated:
str_clean(lmb, "magic/[{vowels}]{2,} => _")
```

*Technically*, the algorithm does not interpolate curly brackets having a 
regular expression meaning. The expression of the form `"{a, b}"` with `"a"` and `"b"`
digits means a repetition of the previous symbol of at least `"a"` times and at most `"b"` times.
The variables are fetched in the calling environment. To obtain fetch them from a different location,
you can use the argument `envir`.

### total {#flag_total}

The flag `"total"` is only available to functions performing a replacement. 
In that case, if a pattern is detected, *the full character string* is replaced 
(instead of just the pattern). 

Ex: let's replace a few car models.
```{r}
cars_small = head(row.names(mtcars))
print(cars_small)

str_clean(cars_small, "ti/mazda => Mazda: sold out!")
```

On top of this, the `"total"` flag allows to perform logical operations across
several regex patterns. You have more information on this in the [dedicated section](#detect_logic).
In a nutshell, you can write `"pat1 & !pat2 | pat3"` with `"patx"` regular 
expresion patterns. This means: contains `pat1` and does not contain `pat2`,
or contains `pat3`.

Ex: detect car brands with a digit and no 'e'.
```{r}
cars_small = head(row.names(mtcars))
print(cars_small)

str_clean(cars_small, "ti/\\d & !e => I don't like that brand!")
```

*Technically*, instead of using `gsub` to replace the pattern, [`str_is`](#detect_funs) is used
to detect which element contains the pattern. Each element with the pattern
is then substituted with the replacement.

### single {#flag_single}

The flag `"single"` is only available to functions performing a replacement.
It allows only a single substitution to take place. Said differently, only the 
first replacement is performed.

Ex: single substitutions.
```{r}
encounter = str_vec("Hi Cyclops., Hi you. What's your name?, Odysseus is my name.")
# we only remove the first word
str_clean(encounter, "single/\\w+ => ...")
```

*Technically*, the function `base::sub` is used instead of `base::gsub`.

## Escaping flags: How to, and a word of caution with paths {#flag_escaping}

If your regular expression contains a slash (`"/"`), this will come in conflict
with the parsing of the optional flags. 
At the moment a `/` is present in a pattern, the algorithm will throw an error
if the expected flags are not written correctly.

To use a slash in the regex without adding flags there are two solutions:

- start the pattern with a `"/"`. This is considered the empty flag and will be discared.
- escape the first `"/"` with a double backslash

Ex: let's invert the numerator and denominator of a division.
```{r}
eq = "5/x = 3/2"
# a) escaping with empty flags
str_clean(eq, "/(\\d)/(\\w) => \\2/\\1")

# b) escaping with backslashes
str_clean(eq, "(\\d)\\/(\\w) => \\2/\\1")
```

**Warning:** when applying regular expressions on file paths, the escaping 
of the slash can be highly confusing. This is a limitation of the current implementation 
of the optional flags. For example `"/[^/]+$"` looks like a legit regex to get 
the parent folder, but it won't work as expected within `stringmagick`. 
To behave as expected, you would need to write `"//[^/]+$"` which looks weird, 
or `"\\/[^/]+$"`.

# Intuivite detection of regex patterns

Detecting a single regex pattern is pretty straightforward with regular tools
like `base::grepl` or `stringr::str_detect`. Things become more complicated
when we want to detect the presence of multiple patterns. 

`stringmagick` offers three functions with an intuitive syntax to deal with 
complex pattern detection:

- [str_is](#detect_funs): returns a logical vector
- [str_which](#detect_funs): returns the indexes of the matches
- [str_get](#detect_funs): returns the matches

As described in the next section., these tools benefit from a streamlined syntax to 
logically combine several regex patterns. 

## Logically combining regex patterns {#detect_logic}

Assume `"pat1"` and `"pat2"` are two regular expression patterns and we want 
to test whether the string `x` contains a combination of these patterns. 
Then:

- `"pat1 & pat2"` = `x` contains `pat1` AND `x` contains `pat2`
- `"pat1 | pat2"` = `x` contains `pat1` OR `x` contains `pat2`
- `"!pat1"` = `x` does not contain `pat1`
- `"!pat1 & pat2"` = `x` does not contain `pat1` AND `x` contains `pat2`

Hence the three logial operators are:

- `" & "`: logical AND, it **must** be a space + an ampersand + a space 
(just the `&` does not work)
- `" | "`: logical OR, it **must** be a space + a pipe + a space 
(just the `|` does not work)
- `"!"`: logical NOT, it works only when it is the first character of the pattern.
Note that anything after it (including spaces and other `!`) *is part of the regular expression*

The parsing of the logical elements is done before any regex interpretation. 
The logical evaluations are done from left to right and are sequentially combined. 

You **cannot** combine logical statements with parentheses. 

For example: `"hello | (world & my lady)"` leads to: `x` contains `"hello"` or contains `"(world"`, 
and contains `"my lady)"`. The two latter are invalid regexes but can make sense if you
have the flag "fixed" turned on. To escape the meaning of the logical operators, 
see the [dedicated section](#logical_escape).

In general, the logical `"not"` always apply to a single pattern and **not** to the full pattern.
But see the [section on flags](#logical_flags) for an exception.

### Escaping the meaning of the logical operators {#logical_escape}

To escape the meaning of the logical operators, there are two solutions to 
escape them:

- use two backslashes just before the operator: `"a \\& b"` means `x`
contains `"a & b"`
- use a regex hack: the previous example is equivalent to `"a [&] b"` in regex parlance
and won't be parsed as a logical AND

### How do regex flags work with logically combined regexes? {#logical_flags}

When you add flags to a pattern, these apply to *all* regex sub-patterns. 
This means that `"f/( | )"` treats the two parentheses as "fixed". 
*You cannot add flags specific to a single sub-pattern.*

The functions `str_is`, `str_which` and `str_get` accept a negation before the 
flags (they're the only ones).   
In that case, the result of all the sub-patterns is negated. You can add the
empty flag to make this work without flags. For example `"!/a & b & c"` means `x` does not contain
(`a` AND `b` AND `c`). 

## Pattern detection with `str_is`, `str_which` and `str_get` {#detect_funs}

Use `str_is`, `str_which` and `str_get` to detect patterns in character vectors and obtain either 
a logical vector, an integer vector, or the values.

In this section we give examples for `str_get` which hopefully will be explicit 
enough to illustrate how it works. For the record, `str_get` uses `str_is` internally so
these examples are equivalent with `str_is` or `str_which`. 

Ex.1: series of examples using the *recommended syntax*.
```{r}
cars = row.names(mtcars)
smagick("All cars from mtcars:\n{C, 60 width ? cars}")

# cars with an 'a', an 'e', an 'i', and an 'o', all in lower case
str_get(cars, "a & e & i & o")

# cars with no 'e' and at least one digit
str_get(cars, "!e & \\d")

# we use the empty flag trick to negate the full pattern
# NOT(cars with an 'e' AND one digit)
str_get(cars, "!/e & \\d")

# flags apply to all
# contains the 'words' 2, 9 or l
# alternative syntax for flags: "wi/2 | 9 | l"
str_get(cars, "word, ignore/2 | 9 | l")
```

The default syntax is `str_get(x, ...)` (same for `str_is` and `str_which`), 
where `...` contains any number of patterns
to detect. By default the results of these pattern detections are combined with 
a logical AND. To combine them with a logical OR, you need to use the argument
`or = TRUE`. You can also pass the flags as regular function arguments. They then
apply to all patterns.

Ex.2: replication of Ex.1 using an alternative syntax.
```{r}
# str_get(cars, "a & e & i & o")
# cars with an 'a', an 'e', an 'i', and an 'o', all in lower case
str_get(cars, "a", "e", "i", "o")

# str_get(cars, "!e & \\d")
# cars with no 'e' and at least one digit
str_get(cars, "!e", "\\d")

# str_get(cars, "!/e & \\d")
# This example cannot be replicated directly, we need to apply logical equivalence
str_get(cars, "!e", "!\\d", or = TRUE)

# str_get(cars, "wi/2 | 9 | l")
# contains the 'words' 2, 9 or l
str_get(cars, "2", "9", "l", or = TRUE, word = TRUE, ignore.case = TRUE)
```

### Specificities of `srt_get` {#detect_get}

On top of the detection previously described, the function `srt_get` changes its 
behavior with the arguments `seq` or `seq.unik`. It also supports [automatic caching](#get_caching).

#### Sequentially appending results

As seen previously, patterns in `...` are combined with a logical AND. If you set
`seq = TRUE`, this behavior changes. The results of each pattern becomes stacked 
sequentially. Schematically, you obtain the vector `c(x_that_contains_pat1, x_that_contains_pat2, etc)`
with `pat1` the first pattern in `...`, `pat2` the second pattern, etc.

Using `seq.unik = TRUE` is like `seq` but applies the function `unique()` at the end.

Ex: sequentially combining results.
```{r}
# cars without digits, then cars with 2 'a's or 2 'e's and a digit
str_get(cars, "!\\d", "i/a.+a | e.+e & \\d", seq = TRUE)

# let's get the first word of each car name
car_first = str_ops(cars, "extract.first")
# we select car brands ending with 'a', then ending with 'i'
str_get(car_first, "a$", "i$", seq = TRUE)
# seq.unik is similar to seq but applies unique()
str_get(car_first, "a$", "i$", seq.unik = TRUE)
```

#### Caching {#get_caching}

At the exploration stage, we often run the same command with a few
variations on the same data set. Acknowledging this, `str_get` supports the caching
of the data argument in interactive use. This means that the user can concentrate in
the pattern to find and need not bother to write the data from where to fectch the
values. Note that `str_get` is the only `stringmagick` function to have this
ability.

Caching is always enabled, you don't need to do anything.

Ex: caching of the data.
```{r}
# Since we used `car_first` in the previous example, we don't need to provide
# it explicitly now
# => brands containing 'M' and ending with 'a' or 'i'; brands containing 'M'
str_get("M & [ai]$", "M", seq.unik = TRUE)
```

# Chaining string operations with `str_ops` {#sec_ops}

Formatting text data often requires applying many functions
(be it for parsing, text analysis, etc). 
Even for simple tasks, the number of operations can quickly balloon, adding many
lines of code, reducing readability, and all this for basic processing.

The function `str_ops` tries to solve this problem. 
It has access to all (50+) [`smagick` operations](https://lrberge.github.io/stringmagick/articles/smagick_walkthrough.html#sec_operations_reference), allowing for
a compact and readable way to chain basic operations on character strings.

Below are a few motivating examples.

Ex.1: Parsing data.
```{r}
# parsing an input: extracting the numbers
input = "8.5in, 5.5, .5 cm"
str_ops(input, "','s, tws, '^\\. => 0.'r, '^\\D+|\\D+$'r, num")
# Explanation------------------------------------------------------------------|
# ','s: splitting w.r.t. ','                                                   |
# tws: trimming the whitespaces                                                |
# '^\\. => 0.'r: adds a 0 to strings starting with '.'                         |
# '^\\D+|\\D+$'r: removes non-digits on both ends of the string                |
# num: converts to numeric                                                     |

# now extracting the units
str_ops(input, "','s, '^[ \\d.]+'r, tws")
# Explanation------------------------------------------------------------------|
# ','s: splitting w.r.t. ','                                                   |
# '^[ \\d.]+'r: removes the ' ', digit and '.' at the beginning of the string  |
# tws: trimming the whitespaces                                                |
```

Ex.2: extracing information from text.
```{r}
# Now using the car data
cars = row.names(mtcars)

# let's get the brands starting with an "m"
str_ops(cars, "'i/^m'get, x, unik")
# Explanation------------------------------------------------------------------|
# 'i/^m'get: keeps only the elements starting with an m,                       |
#            i/ is the 'regex-flag' "ignore" to ignore the case                |
#            ^m means "starts with an m" in regex language                     |
# x: extracts the first pattern. The default pattern is "[[:alnum:]]+"         |
#    which means an alpha-numeric word                                         |
# unik: applies unique() to the vector                                         |

# let's get the 3 largest numbers appearing in the car models
str_ops(cars, "'\\d+'x, rm, unik, num, dsort, 3 first")
# Explanation------------------------------------------------------------------|
# '\\d+'x: extracts the first pattern, the pattern meaning "a succession"      |
#          of digits in regex language                                         |
# rm: removes elements equal to the empty string (default behavior)            |
# unik: applies unique() to the vector                                         |
# num: converts to numeric                                                     |
# dsort: sorts in decreasing order                                             |
# 3 first: keeps only the first three element                                  |
```

As you can see, an operation that would take multiple lines to read and understand
now can be read from left to right in a single line.

# `str_clean`: One function to clean them all {#sec_clean}

The function `str_clean` streamlines the cleaning of character vectors by providing:

- i) a specialized syntax to replace multiple regex patterns, 
- ii) a direct access to many low level string operations, and 
- iii) the ability to chain these two operations. 

## Cleaning syntax

This function is of the form `str_clean(x, ...)` with `x` the vector to clean and
`...` any number of cleaning operations which can be of two types:

1. use `"pat1, pat2 => replacement"` to replace the regex patterns `pat1` and `pat2` 
with the value `replacement`.
1. use `"@op1, op2"` to perform any arbitrary sequence of [`smagick` operation](https://lrberge.github.io/stringmagick/articles/smagick_walkthrough.html#sec_operations_reference)

In the operation `"pat1, pat2 => replacement"`, the pattern is first split with respect
to the pipe, `" => "` (change it with argument `pipe`), to get `replacement`. 
Then the pattern is split with respect to commas (i.e. `",[ \t\n]+"`, change it with argument `sep`) 
to get `pat1` and `pat2`. A sequence of `base::gsub` calls is performed to replace each
`patx` with `replacement`. 

By default the replacement is the empty string. This means that writting `"pat1, pat2"`
will lead to erasing these two patterns.

If a pattern starts with an `"@"`, the subsequent character string is sent to `str_ops`.
For example `"@ascii, lower"` is equivalent to `str_ops(x, "ascii, lower")` which 
turns `x` to ASCII and lowers the case. 

## Example of text cleaning {#clean_example}

```{r}
monologue = c("For who would bear the whips and scorns of time",
              "Th' oppressor's wrong, the proud man's contumely,",
              "The pangs of despis'd love, the law's delay,",
              "The insolence of office, and the spurns",
              "That patient merit of th' unworthy takes,",
              "When he himself might his quietus make",
              "With a bare bodkin? Who would these fardels bear,",
              "To grunt and sweat under a weary life,",
              "But that the dread of something after death-",
              "The undiscover'd country, from whose bourn",
              "No traveller returns- puzzles the will,",
              "And makes us rather bear those ills we have",
              "Than fly to others that we know not of?")

# Cleaning a text
str_clean(monologue, 
          # use smagick to: lower the case and remove basic stopwords
          "@lower, stopword",
          # remove a few extra stopwords(we use the flag word 'w/')
          "w/th, 's",
          # manually stem some verbs
          "despis'd => despise", "undiscover'd => undiscover", "(m|t)akes => \\1ake",
          # still stemming: dropping the ending 's' for words of 4+ letters, except for quietus
          "(\\w{3,}[^u])s\\b => \\1",
          # normalizing the whitespaces + removing punctuation
          "@ws.punct")

```


