---
title: "`stringmagic`: A quick tour"
author: "Laurent R. Bergé"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: journal
    highlight: haddock
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
      smooth_scroll: no
  pdf_document:
    toc: yes
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{stringmagic_tour}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

# we preload it to avoid ugly (was comiled with R.x.x) warnings in the doc
library(stringmagic)

options("smagic_str_get_forced_caching" = TRUE)
```

Text is everywhere in data science, so better be equipped with the right tools!
The package `stringmagic` offers several functions to empower and streamline low level 
operations on text data.

This short vignette provides a quick tour of `stringmagic` tools 
and links to the associated documentations.

# String interpolation with `smagic`

The function `smagic` is the *raison d'être* of this package. It offers *magical*
string interpolation.

For the record, string interpolation is a feature common to many programming languages, 
it is about inserting values from a variable directly in a character string.
For example if `x = "Sam"`, then `"Hi {x}!"` becomes "Hi Sam!".

`smagic` enables to perform *any arbitrary operation* to the variables we interpolate. 
This opens the door to many possibilities, and makes the composition
of complex character strings a piece of cake. 
By default over 50 basic operations are available, but the user can seamlessly add any she wants.
It supports nesting, has advanced support for pluralization, and much more. 

Like any powerful tool (looking at you *regular expressions*), it has its own syntax,
its own commands, and will require practice before mastering it. To minimize learning costs, 
great effort has been spent on error handling, with error
messages hopefully explaining where the problems are and how to solve them. So fear not 
trying out stuff!

If interested, here is the [documentation](https://lrberge.github.io/stringmagic/articles/smagic_walkthrough.html) and next comes an example. 

## Example

Here's a set of stuff I do with my kids and my associated self-assessed sikll level:

```{r}
skills = c("managing to lose at chance-based board games: 8/10", 
           "making them clean their mess: 3/10", 
           "being OK with cooking stuff they don't eat: 9/10", 
           "navigating the grocery store without tears: 5/10",
           "maximizing swing push strength while (mostly) avoiding falls: 7/10")
```

Now say I want to write a small message and then display the first `n` of those skills
by decreasing skill order, with `n` an arbitrary number. Here's how it could be done:
```{r}
library(stringmagic)
n = 3
smagic("My {&n>1;{N?.} }best skill{#s} with kids:\n",
        "{'^\\D+'dsort, `n`first, ' - 'paste, '\n'c ? skills}")

# with another number
n = 1
smagic("My {&n>1;{N?.} }best skill{#s} with kids:\n",
        "{'^\\D+'dsort, `n`first, ' - 'paste, '\n'c ? skills}")
```

To understand how this works, please have a look at the [dedicated vignette](https://lrberge.github.io/stringmagic/articles/smagic_walkthrough.html).

Even if it looks esoteric at first sight, writing this kind of dynamic messages is very common,
especially when developping packages.

**Tip:** One way to figure out the value added of `smagic` is trying to write this (seemingly simple) 
message with alternative tools. 

# Other string tools

On top of `smagic`, this package offers a few functions providing new ways to tackle old problems:

- [how to](#fun_clean) efficiently **and** *explicitly* perform manifold string cleaning operations
- [how to](#fun_detect) detect the presence of several regular expressions at once
- [how to](#fun_chain) chain basic string operations without exploding the number of lines of code

Here we provide an overview, for more information, see the [associated vignette](https://lrberge.github.io/stringmagic/articles/string_tools.html).

## Cleaning text: Quick and easy please {#fun_clean}

One common (and tedious) task in data sciencce is *cleaning* character strings. 
The function `str_clean` is specialized in cleaning and has a dedicated syntax:

- use `"pat1, pat2 => r"` to replace the regex patterns `pat1` and `pat2` with the value `r`.
- use `"@op1, op2"` to use `smagic` operations.

To illustrate this function, since real data science problems are a tiny bit dull, 
allow me to use a French classic:

```{r}
tragedy = c("P: Grâces au ciel, mes mains ne sont point criminelles.",
            "Plût aux dieux que mon coeur fût innocent comme elles !", 
            "...",
            "P: Quel fruit espères-tu de tant de violence ?",
            "Tu frémiras d'horreur si je romps le silence.",
            "O: Et que me direz-vous, qui ne cède, grands dieux !",
            "À l'horreur de vous voir expirer à mes yeux ?",
            "P: Quand tu sauras mon crime, et le sort qui m'accable,",
            "Je n'en mourrai pas moins, j'en mourrai plus coupable.")

# Cleaning a text
str_clean(tragedy, 
          # use smagic to: turn to ascii with transliteration, lower the case
          # remove the punctuation + isolated letters + normalize the whitespaces, 
          # remove empty strings
          "@ascii, lower, ws.punct.isolated, rm",
          # remove a few French stopwords (I use the 'word' flag with w/)
          "w/au, aux, mes?, que, mon, comme, elles?, quel, tu, de, si, je, le, et, qui, en",
          # manually stem some verbs
          "fremiras => fremir", "w/sont, fut => etre", "w/plut => plaire", "mourrai => mourir",
          # still stemming, note that you can use regex groupings as usual
          "romps => rompre", "sauras => savoir", "\\b(accable|cede|espere)s? => \\1r",
          # renormalizing the whitespaces
          "@ws")
```

## Fast character string detection {#fun_detect}

When you want to detect a single pattern in a string, `base::grepl` or 
`stringr::str_detect` works pretty well. You feed a regex, and done. 
But what if you want to detect the presence, at the same time, of two distinct regular expressions.
This quickly becomes messier (try to get the examples below with a single `grepl` call!).

For pattern detection, `stringmagic` offers three functions: `str_is`, `str_which` and `str_get`.
Their syntax is as follows: `"pat1 & !pat2 | pat3"` with `patx` a regular expression. 
This example means: find the elements that: contain `pat1` and do not contain `pat2`, 
or contain `pat3`. See the examples below:

```{r}
cars = row.names(mtcars)
# Looking at the cars
smagic("Cars from the `mtcars` data set:\n{sort, C, 80 width ? cars}.")

# cars containing 'a', 'e', 'i' and 'o'
str_get(cars, "a & e & i & o")

# in interactive mode, str_get keeps the previous item in memory, 
# so you can avoid providing the first argument

# cars not containing any digit nor the letter 'e'
str_get("!\\d & !e")
```

## Chaining string operations {#fun_chain}

If you want to apply a sequence of operations to a single character vector,
you can use `str_ops`. It has access to all operations available to `smagic`.

```{r}
cars = row.names(mtcars)
# Looking at the cars
smagic("Cars from the `mtcars` data set:\n{sort, C, 80 width ? cars}.")

# let's get the brands starting with an "m"
str_ops(cars, "'i/^m'get, x, unik")

# let's get the 3 largest numbers appearing in the car models
str_ops(cars, "'\\d+'x, rm, unik, num, dsort, 3 first")
```



